### 2022-10-20 manuscript/manuscript.tex
theory to main result

\section*{Introduction}

As genomes are transmitted through the generations, lifetime events selectively
alter what is passed to the next generation. While selection directly acts on
the genetic variants that cause heritable fitness differences among
individuals, neighboring regions experience an indirect contraction in genetic
variability. This byproduct is known as \emph{linked selection}, and it occurs
when selection changes a site's frequency on timescale faster than
recombination can decouple that site from its genetic background. Because
linked selection modifies the genetic variability in a regionally-correlated
way, it creates observable patterns in genomic data that are used to understand
the type of selection processes acting on genomes and estimate key evolutionary
quantities.

@@

Much work has been done on developing mathematical theory for two linked
selection processes: hitchhiking and background selection. These are
distinguished by the evolutionary processes that generate fitness variation and
the resulting selection dynamics. Under the classic hitchhiking model, a new
beneficial mutation sweeps fixation, leading to a dip in genetic diversity on a
small spatial scale. Since the sweeps leave relatively narrow footprint, they
influence genome-wide diversity only when beneficial mutations are recurring
and recombination is relatively limited, as in \emph{Drosophila}. These models
assume sweeps are sufficiently rare as to not occur simultaneously, which is
consistent with the slow estimated rate of strongly beneficial mutations (XXX).
Overall, the classic hitchhiking model is simple because the underlying
selection process does not generate polygenic fitness variation, even though
this is type of variation plays a dominant role in adaptation (XXX).

By contrast, background selection is a consequence of the fitness variation
created by a steady influx of new deleterious mutations. The classic background
selection model (XXX) considers how the balance between mutation introducing
deleterious variants into the population and selection removing them, reduces
nearby genetic variability. A critical assumption of this model is that such an
equilibrium between mutation and selection exists, which is the case only when
these mutations have a large enough fitness cost to ensure that they cannot fix
in the population. Under this process, the number of deleterious mutations per
gamete reaches a Poisson stationary distribution with rate $\nicefrac{U}{s}$,
where $U = \mu_D L$ is the deleterious mutation rate ($\mu_D$ per basepair per
generation) in a region of $L$ basepairs for mutations with a heterozygous
selection coefficient of $s$ (XXX). The fraction of gametes without any
mutations is then $f_0 = \exp(-\nicefrac{\mu_D}{s})$. Without recombination,
any gamete with deleterious mutations is destined to bring its linked neutral
variation with it to extinction, effectively reducing diversity levels to those
expected in a neutrally-evolving population of size $f_0 N$.

Remarkably, predictions from simple background selection models with
recombination closely match large-scale patterns of human diversity and
divergence (XXX). Since background approximately rescales a neutral coalescent
process by $f_0$, the predicted diversity is reduced from neutral levels by a
factor $B = \nicefrac{N_e}{N}$, which is the ratio of effective population size
under background selection to the population size. In background selection
models with recombination, $B$ is determined by the number of conserved
basepairs and recombination in a region. Consequently the predicted reduction
along the genome (known as $B$ maps) mirror the regional density of conserved
elements under purifying selection and recombination rates. These $B$ maps have
been used in studies of XXX, are inputs into machine learning models that
predict functional constraint, and XXX. Furthermore, since new deleterious
mutations are perpetually flowing into populations $B$ maps may provide a
better null model for evolution than the standard neutral model.

While background selection models do predict diversity levels  XXX

% outline?
- Disentangling evolutionary processes from the patterns of genetic variation
across the genome. The structure of what can be under selection shapes these
patterns when selection acts; when drift acts, these patterns should not exist.

- For species with small effective population sizes (and perhaps large map
lengths), BGS seems to be a dominant selective force in shaping patterns of
diversity (e.g. at the megabase scale in humans).

- However, current approaches to fitting BGS models fail under the weak
selection domain. This weak selection domain is important (discuss
HRI/ratchet/etc). Importantly, the characteristic scale of the impact of linked
selection is $\nicefrac{s}{r}$, meaning weak selection acts on smaller scales
than strong selection. Our ability to see selection at small scales may be
inhibited by the fact that our current BGS theory only works for strong
selection.

- We need to say briefly that we're using BGS to mean linked selection due to
purifying selection, and thus the strong and weak regimes (classically it has
just meant the strong regime).

\section*{Conclusion}

- care should be taken applying these methods to \emph{Drosophila}, since the
$\delta_{LD}$ term could be much larger.

\section*{Results}

\subsection*{Theory}

- overview of why BGS models fail?

From an evolutionary perspective, linked selection is important because by
modifying neighboring levels of genetic diversity, it is also directly altering
the selection dynamics at these neighboring sites.

The other use of these models is to understand how linked selection processes
differentially alter the genetic variability, or equivalently, the genealogical
trees, along the genome.

XXX how many sites are under selection at any moment? What's fitness variation?

Early on, some pointed out that evidence of strong selective sweeps in the
human genome was not the full story

Generally, in organisms with large amounts of recombination, linked selection
has

Weak selection processes

distorts
genome-wide

diversity.

The linked selection literature has primarily focused on developing theory for,
and quantifying the effects of two linked selection processes: hitchhiking and
background selection. Genome-wide models of the hitchhiking process consider
how beneficial mutations that sporadically enter the population

- QGLS polygenic linked selection

- H and W

- inflation factor

\subsection*{Statistical Methods}

@@

Since the probability of coalesce (or, equivalently, identity by descent) is
proportional to the variance in allele frequency change
\parencite{Barton2000-zg}, this encodes the pairwise coalescent rate of the
population through time under selection through the $N_{e,t}$ terms.  In the
original derivation of \textcite{Santiago1995-hx}, they set each $N_{e,t}$ term
to a single $N$ and solve this summation in the limit to infinity giving,

@@

\begin{align}
  N_{e,\infty}^* \approx N \exp\left(-\frac{V_h}{(1-Z) (2-(2-L) Z)} \right).
\end{align}

@@

Under strong background selection, multiple segregating deleterious sites
contribute to fitness variation. BGS models assume a multiplicative fitness
model, such that the fitness of an individual $i$ is

\begin{align}
  w_i = (1-s)^{n_i}
\end{align}

where $n_i$ is the number of deleterious mutations this individual carries.

Classic background selection theory assumes an infinite population size, such
that mutation frequencies are at their mutation-selection equilibrium and the
expected number of mutations across individuals is $\E_i(n_i) =
\nicefrac{\mu}{s}L$ where $L$ is the total number of basepairs that can be
mutated (the mutational target size).  Under a multiplicative fitness function
in an infinite population, selection cannot generate linkage disequilibria
\parencite{Turelli1990-kd}. Thus the classic strong background selection model
assumes additive genetic fitness is equal to the additive genic fitness (which
excludes the contribution of covariance between selected sites),

\begin{align}
  V_\text{BGS} &:= \V_i(w_i) = 2 \sum_{l=1}^L p(1-p) s^2 \\
            &= 2 L \frac{\mu}{s} \left( 1 - \frac{\mu}{s}\right) s^2 = U s + \mathcal{O}(\mu^2) \approx Us.
\end{align}

where $U$ is defined $U := 2 L \mu$, the deleterious mutation rate per diploid
genome, per generation. They then set $V_A = V_\text{BGS}$, and note that $V_m
\approx Us^2$, since mutations have frequency of $\nicefrac{1}{2N}$ and each
increase the variance by a factor of $s^2$. Note that because the classic BGS
model assumes deterministic selection dynamics (i.e. an infinite population),
the equilibrium $Z$ given by equation \eqref{eq:Z} excludes the reduction due
to drift. The reduction factor under the strong BGS model is,

@@

Using $Z_\text{BGS}$, $V_m$, and $V$ under the background selection model, and
approximation $N_{e,\infty}^\text{SC98}$ (equation XXX), find an equation
reminiscent of the classic background selection with recombination results. We
find that by using the full equation XXX $N_{e,\infty}^*$ rather than the
approximation of Santiago and Caballero, the reduction due to background
selection is

@@

which matches equation 8 of \textcite{Hudson1995-xc} up to the $Ls$ term and
equation 10 of \textcite{Nordborg1996-nq} up to the $2s$ term.

\subsection*{Weak Selection Dynamics}

With the possibility of weak selection, deleterious alleles can drift up in
frequency and fix. Consequently, the fitness distribution is no longer
stationary, as the set of haplotypes without any deleterious mutations can be
lost due to drift. The approach of \textcite{Santiago2016-mu} that we implement
in our methods is to derive an equation for

@@

\subsection*{The Rate of the Ratchet}

\section*{Methods}

\subsection*{Data Preprocessing}

\subsection*{Human}

Site-level neutral and accessibility masks

Bin-level masks based on fraction of accessible bases in a window (7\%).

Bin-level tail filtering.

>>>>

% \section*{Introduction}

% As genomes are transmitted through the generations, lifetime events selectively
% alter what is passed to the next generation. While selection directly acts on
% the genetic variants that cause heritable fitness differences among
% individuals, neighboring regions experience an indirect contraction in genetic
% variability. This byproduct is known as \emph{linked selection}, and it occurs
% when selection changes a site's frequency on timescale faster than
% recombination can decouple that site from its genetic background. Because
% linked selection modifies the genetic variability in a regionally-correlated
% way, it creates observable patterns in genomic data that are used to understand
% the type of selection processes acting on genomes and estimate key evolutionary
% quantities.

@@

% Much work has been done on developing mathematical theory for two linked
% selection processes: hitchhiking and background selection. These are
% distinguished by the evolutionary processes that generate fitness variation and
% the resulting selection dynamics. Under the classic hitchhiking model, a new
% beneficial mutation sweeps fixation, leading to a dip in genetic diversity on a
% small spatial scale. Since the sweeps leave relatively narrow footprint, they
% influence genome-wide diversity only when beneficial mutations are recurring
% and recombination is relatively limited, as in \emph{Drosophila}. These models
% assume sweeps are sufficiently rare as to not occur simultaneously, which is
% consistent with the slow estimated rate of strongly beneficial mutations (XXX).
% Overall, the classic hitchhiking model is simple because the underlying
% selection process does not generate polygenic fitness variation, even though
% this is type of variation plays a dominant role in adaptation (XXX).

% By contrast, background selection is a consequence of the fitness variation
% created by a steady influx of new deleterious mutations. The classic background
% selection model (XXX) considers how the balance between mutation introducing
% deleterious variants into the population and selection removing them, reduces
% nearby genetic variability. A critical assumption of this model is that such an
% equilibrium between mutation and selection exists, which is the case only when
% these mutations have a large enough fitness cost to ensure that they cannot fix
% in the population. Under this process, the number of deleterious mutations per
% gamete reaches a Poisson stationary distribution with rate $\nicefrac{U}{s}$,
% where $U = \mu_D L$ is the deleterious mutation rate ($\mu_D$ per basepair per
% generation) in a region of $L$ basepairs for mutations with a heterozygous
% selection coefficient of $s$ (XXX). The fraction of gametes without any
% mutations is then $f_0 = \exp(-\nicefrac{\mu_D}{s})$. Without recombination,
% any gamete with deleterious mutations is destined to bring its linked neutral
% variation with it to extinction, effectively reducing diversity levels to those
% expected in a neutrally-evolving population of size $f_0 N$.

% Remarkably, predictions from simple background selection models with
% recombination closely match large-scale patterns of human diversity and
% divergence (XXX). Since background approximately rescales a neutral coalescent
% process by $f_0$, the predicted diversity is reduced from neutral levels by a
% factor $B = \nicefrac{N_e}{N}$, which is the ratio of effective population size
% under background selection to the population size. In background selection
% models with recombination, $B$ is determined by the number of conserved
% basepairs and recombination in a region. Consequently the predicted reduction
% along the genome (known as $B$ maps) mirror the regional density of conserved
% elements under purifying selection and recombination rates. These $B$ maps have
% been used in studies of XXX, are inputs into machine learning models that
% predict functional constraint, and XXX. Furthermore, since new deleterious
% mutations are perpetually flowing into populations $B$ maps may provide a
% better null model for evolution than the standard neutral model.

% While background selection models do predict diversity levels  XXX

% % outline?
% - Disentangling evolutionary processes from the patterns of genetic variation
% across the genome. The structure of what can be under selection shapes these
% patterns when selection acts; when drift acts, these patterns should not exist.

% - For species with small effective population sizes (and perhaps large map
% lengths), BGS seems to be a dominant selective force in shaping patterns of
% diversity (e.g. at the megabase scale in humans).

% - However, current approaches to fitting BGS models fail under the weak
% selection domain. This weak selection domain is important (discuss
% HRI/ratchet/etc). Importantly, the characteristic scale of the impact of linked
% selection is $\nicefrac{s}{r}$, meaning weak selection acts on smaller scales
% than strong selection. Our ability to see selection at small scales may be
% inhibited by the fact that our current BGS theory only works for strong
% selection.

% - We need to say briefly that we're using BGS to mean linked selection due to
% purifying selection, and thus the strong and weak regimes (classically it has
% just meant the strong regime).

% \section*{Conclusion}

% - care should be taken applying these methods to \emph{Drosophila}, since the
% $\delta_{LD}$ term could be much larger.

% \section*{Results}

% \subsection*{Theory}

% - overview of why BGS models fail?

% From an evolutionary perspective, linked selection is important because by
% modifying neighboring levels of genetic diversity, it is also directly altering
% the selection dynamics at these neighboring sites.

% The other use of these models is to understand how linked selection processes
% differentially alter the genetic variability, or equivalently, the genealogical
% trees, along the genome.

% XXX how many sites are under selection at any moment? What's fitness variation?

% Early on, some pointed out that evidence of strong selective sweeps in the
% human genome was not the full story

% Generally, in organisms with large amounts of recombination, linked selection
% has

% Weak selection processes

% distorts
% genome-wide

% diversity.

% The linked selection literature has primarily focused on developing theory for,
% and quantifying the effects of two linked selection processes: hitchhiking and
% background selection. Genome-wide models of the hitchhiking process consider
% how beneficial mutations that sporadically enter the population

% - QGLS polygenic linked selection

% - H and W

% - inflation factor

% \subsection*{Statistical Methods}

@@

Since the probability of coalesce (or, equivalently, identity by descent) is
proportional to the variance in allele frequency change
\parencite{Barton2000-zg}, this encodes the pairwise coalescent rate of the
population through time under selection through the $N_{e,t}$ terms. Others
have found that the genealogies under purifying selection can be characterized
by such a time-dependent effective population size
\parencite{Nicolaisen2013-gv}. In the original derivation of
\textcite{Santiago1995-hx}, they set each $N_{e,t}$ term to a single $N$ and
solve this summation in the limit to infinity giving,

@@

\begin{align}
  N_{e,\infty} \approx N \exp\left(-\frac{V_h}{(1-Z) (2-(2-L) Z)} \right).
\end{align}

@@

Under strong background selection, multiple segregating deleterious sites
contribute to fitness variation. BGS models assume a multiplicative fitness
model, such that the fitness of an individual is

\begin{align}
  w = (1-s)^{n}
\end{align}

where $n$ is the number of deleterious mutations this individual carries.

Classic background selection theory assumes an infinite population size, such
that mutation frequencies are at their mutation-selection equilibrium and the
expected number of mutations across individuals is $\E(n) = \nicefrac{\mu}{s}L$
where $L$ is the total number of basepairs that can be mutated (the mutational
target size).  Under a multiplicative fitness function in an infinite
population, selection cannot generate linkage disequilibria
\parencite{Turelli1990-kd}. Thus the classic strong background selection model
assumes additive genetic fitness is equal to the additive genic fitness (which
excludes the contribution of covariance between selected sites),

\begin{align}
  \label{eq:va_bgs}
  V_\text{BGS} &:= \V(w) = 2 s^2 \sum_{l=1}^L p_l(1-p_l)  \\
            &= 2 L s^2 \frac{\mu}{s} \left( 1 - \frac{\mu}{s}\right) = U s + \mathcal{O}(\mu^2) \approx Us.
\end{align}

since under mutation-selection balance $p_l = \nicefrac{\mu}{s}$ and $U$ is
defined $U := 2 L \mu$, the deleterious mutation rate per diploid genome, per
generation. They then set $V_A = V_\text{BGS}$, and note that $V_m \approx
Us^2$, since mutations have frequency of $\nicefrac{1}{2N}$ and each increase
the variance by a factor of $s^2$. Note that because the classic BGS model
assumes deterministic selection dynamics (i.e. an infinite population), the
equilibrium $Z$ given by equation \eqref{eq:Z} excludes the reduction due to
drift. The reduction factor under the strong BGS model is,

@@

Using $Z_\text{BGS}$, $V_m$, and $V$ under the background selection model, and
approximation $N_{e,\infty}^\text{SC98}$ (equation XXX), find an equation
reminiscent of the classic background selection with recombination results. We
find that by using the full equation XXX $N_{e,\infty}$ rather than the
approximation of Santiago and Caballero, the reduction due to background
selection is

@@

which matches equation 8 of \textcite{Hudson1995-xc} except for the $Ls$ term
and equation 10 of \textcite{Nordborg1996-nq} except for the $2s$ term.

\subsection*{Effective Population Size Under Weak and Strong BGS}

With the possibility of weak selection, deleterious alleles can drift up in
frequency and fix. Consequently, the fitness distribution is no longer
stationary, as the set of haplotypes without any deleterious mutations can be
lost due to drift. This changes the shape of the fitness distribution, and thus
the variance parameter that our model relies upon
\parencite{Gessler1995-hz,OFallon2010-my,Good2013-lp,Haigh1978-gt,Higgs1995-xc}.
So far, finding an explicit equation for the variance has been difficult due to
the ``moment-closure problem": the variance depends on higher moments such as
the skew, which in turn depends on higher moments. While the weak selection
regime is composed of an interaction of mutation, selection, drift, and linkage
processes and the dynamics that differ qualitatively based on parameter
combinations. In their approximation, \textcite{Santiago2016-mu} consider the
modified additive genic variance due to the fixation of weakly deleterious
mutations. This fixation process is related to the Muller's Ratchet, the rate
$R$ of fixation of deleterious mutations per generation per region, since each
fixation directly reduces the additive genic variance of the fitness
distribution. However, in general, the rate of this ratchet is unknown since it
depends on the fitness variance (and thus higher moments of the fitness
distribution).

In their 2016 paper, Santiago and Caballero derive a general equation for the
fitness variance that accounts for the reduction due to the fixation of weakly
deleterious alleles (i.e. the rate of Muller's ratchet). There derivation for
the variance uses a heuristic approach based on Fisher's Fundamental Theorem of
Natural Selection (equation 2, \cite{Santiago2016-mu}, equation 15A
\cite{Garcia-Dorado2007-jj}). A more formal derivation follows from
\textcite{Higgs1995-xc} (though they assume an asexual system, the result is
the same). In their work, they derive equations for the change per generation
for the moments and cross-moments of the fitness distribution under
multiplicative selection in a haploid population. For a Poisson number of new
mutations per gamete, per generation, they find (equation 13.3,
\cite{Higgs1995-xc}) that the rate of the ratchet is

\begin{align}
  R = U - s \V(n).
\end{align}

The rate of the ratchet for a diploid population is the same, as long as $U$ is
the mutation rate per generation per diploid genome (as is assumed here).
$\V(n)$ is the variance of the distribution of the \emph{number} of deleterious
mutations (not the fitness distribution); as long as selection is not too
strong, $V_A \approx -s^2\V(n)$. Then, the rate of the ratchet in terms of the
fitness variance is,

\begin{align}
  %R &= U - s\V(n) \\
  R &\approx U + \frac{V_A}{s} \\
  V_A &\approx Us - Rs. 
\end{align}

This is identical to their equation 2, noting that the waiting time between
fixations (i.e. ratchet clicks) is $T = \nicefrac{1}{R}$. Note that if selected
sites are sufficiently deleterious that they cannot fix (or the population is
infinite), $R=0$ and the additive fitness variance is identical to that under
strong background selection, $V_A = Us = V_\text{BGS}$ (equation
\eqref{eq:va_bgs}).

Then, the fixation probability of a new deleterious mutation with additive
effects ($1-s$ in the heterozygote, $1-2s$ in the homozygote) in a diploid
population is,

\begin{align}
  p_\text{fix} = \frac{1-\exp(\nicefrac{2 N_e s}{N})}{1-\exp(4 N_e s)} \approx
  \frac{2 N_e s}{N \left(\exp(4 N_e s)-1\right)}
\end{align}

(c.f. \cite{Durrett2008-ql}, equation 7.21; \cite{Kimura1957-rk}, equation
5.6). Then, Santiago and Caballero argue that the rate of the ratchet is the
inverse of the averaging waiting time until a fixation, $T \approx 1/(2UN
p_\text{fix})$, so

\begin{align}
  T \approx \frac{\exp(4 N_e s)-1}{4 U s N_e}.
\end{align}

Since the ratchet rate depends on $N_e$, which in turn depends on the rate of
the ratchet, we write the inverse as,

\begin{align}
  R(N_e) := \frac{4 U s N_e}{\exp(4 N_e s)-1}
\end{align}

Under weak and strong background selection the variance input of mutations
$V_m$ is the same, and using the $V_A$ derived above, the decay in variance is

\begin{align}
  Z_\text{WS} = 1 - \frac{V_m}{V_A} = 1 - \frac{Us^2}{(U-R(N_e))s} = 1 - \frac{Us}{U-R(N_e)}.
\end{align}

At equilibrium, the $N_e$ that determines fixation probability is the
asymptotic $N_{e,\infty}$. Consequently, Santiago and Caballero argue that the
asymptotic reduction in effective population size is given by the following
system of non-linear equations,

\begin{align}
  R(N_{e,\infty}) &= \frac{4 U s N_{e,\infty}}{\exp(4 N_{e,\infty} s)-1}\\
  N_{e,\infty} &= N \exp\left(-\frac{(U-R(N_{e,\infty}))s}{\left(1-\frac{Us}{U-R(N_{e,\infty})}\right) \left(2-(2-L) \frac{Us}{U-R(N_{e,\infty})}\right)} \right).
\end{align}

@@

% \subsection*{The Rate of the Ratchet}

% \section*{Methods}

% \subsection*{Data Preprocessing}

% \subsection*{Human}

% Site-level neutral and accessibility masks

% Bin-level masks based on fraction of accessible bases in a window (7\%).

% Bin-level tail filtering.
